---
title: "NHANES Hypertension Analysis: Uncovering Undiagnosed Risk"
author: "Fuaad Hassan"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    code-fold: show
---

## Executive Summary

This study analyzes nationally representative survey data from the **National Health and Nutrition Examination Survey (NHANES)** to evaluate demographic patterns in hypertension and identify key risk factors associated with remaining **undiagnosed**.

Using proper survey design adjustments (`psu`, `strata`, and `weight_mec`), I evaluate biological blood pressure readings against self-reported diagnoses and fit a weighted logistic regression model.

---

## Data Pipeline & Survey Methodology

1. **Ingest (`01_ingest.R`)**: Downloaded demographic (`DEMO`), blood pressure (`BPX`), and questionnaire (`BPQ`) datasets directly from CDC servers via `nhanesA`.
2. **Clean (`02_clean.R`)**: Standardized blood pressure measurements, filtered for adult records ($\ge 18$), and categorized patients into four distinct hypertension states: *Healthy*, *Managed*, *Uncontrolled*, and *Undiagnosed*.
3. **Descriptive (`03_descriptive.R`)**: Constructed a complex survey design object using `srvyr` and calculated population-weighted demographic summaries.
4. **Inferential (`04_inferential.R`)**: Restricted analysis strictly to the biologically hypertensive subpopulation using domain estimation (preserving full PSU/Strata variance architecture) and ran a survey-weighted logistic regression (`svyglm`).

---

```{r}
#| label: setup-and-data-prep
#| warning: false
#| message: false

library(dplyr)
library(survey)
library(srvyr)
library(broom)
library(ggplot2)
library(knitr)

# 1. Load Processed Data
final_df <- readRDS("data/processed/nhanes_processed.rds")

# 2. Create Age Brackets and Undiagnosed Target Flag
df_prepared <- final_df %>%
  mutate(
    age_bracket = cut(age, breaks = c(17, 39, 59, Inf), labels = c("18-39", "40-59", "60+")),
    undiagnosed_flag = factor(
      ifelse(htn_status == "Undiagnosed", "Undiagnosed", "Diagnosed"),
      levels = c("Diagnosed", "Undiagnosed")
    )
  )

# 3. Initialize Full Survey Design (Preserves PSU/Strata Variance)
full_svy <- df_prepared %>%
  as_survey_design(
    ids = psu,
    strata = strata,
    weights = weight_mec,
    nest = TRUE
  )

# 4. Domain Subpopulation: Biologically Hypertensive Adults Only
htn_svy <- full_svy %>% filter(is_biological_htn == TRUE)

# 5. Fit Survey-Weighted Logistic Regression
model_svy <- svyglm(
  undiagnosed_flag ~ age + gender + income_ratio,
  design = htn_svy,
  family = quasibinomial()
)

model_results <- tidy(model_svy, exponentiate = TRUE, conf.int = TRUE)

```

## Key Results

### 1. Weighted Population Distribution

Survey-weighted proportions across age brackets demonstrate a clear pattern: while older adults ($\ge 60$) have higher overall rates of hypertension, the proportion of **undiagnosed** cases peaks significantly in mid-adulthood ($40-59$).

```{r}
#| label: fig-descriptive
#| fig-cap: "Weighted Proportion of Hypertension Status by Age Group"
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| message: false

# Calculate weighted proportions by age bracket and status
df_plot <- full_svy %>%
  group_by(age_bracket, htn_status) %>%
  summarize(prop = survey_mean(), .groups = "drop")

ggplot(df_plot, aes(x = age_bracket, y = prop, fill = htn_status)) +
  geom_col(position = "fill", color = "black", alpha = 0.85) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c(
      "Healthy"      = "#4CAF50",
      "Managed"      = "#2196F3",
      "Uncontrolled" = "#FF9800",
      "Undiagnosed"  = "#F44336"
    )
  ) +
  theme_minimal() +
  labs(
    title = "True Hypertension Status by Age Group",
    subtitle = "Adjusted using NHANES complex survey weights",
    x = "Age Bracket",
    y = "Weighted Proportion of Population",
    fill = "Status"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

```

---

### 2. Predictors of Remaining Undiagnosed

The table below details the survey-adjusted Odds Ratios for predicting whether a biologically hypertensive individual remains undiagnosed:

```{r}
#| label: model-results-table
model_results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "age"          ~ "Age (per year)",
      term == "genderFemale" ~ "Female (vs. Male)",
      term == "income_ratio" ~ "Poverty-Income Ratio",
      TRUE ~ term
    ),
    p_formatted = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
  ) %>%
  select(
    Predictor = term, 
    `Odds Ratio (OR)` = estimate, 
    `95% CI Lower` = conf.low, 
    `95% CI Upper` = conf.high, 
    `p-value` = p_formatted
  ) %>%
  kable(digits = 3, caption = "Survey-Weighted Logistic Regression Model Results")

```

```{r}
#| label: fig-forest-plot
#| fig-cap: "Forest Plot of Odds Ratios for Undiagnosed Hypertension"
#| fig-width: 8
#| fig-height: 4.5
#| warning: false
#| message: false

plot_data <- model_results %>% 
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "age"          ~ "Age (per year)",
      term == "genderFemale" ~ "Female (vs. Male)",
      term == "income_ratio" ~ "Poverty-Income Ratio",
      TRUE ~ term
    ),
    term = factor(term, levels = c("Poverty-Income Ratio", "Female (vs. Male)", "Age (per year)"))
  )

ggplot(plot_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2, color = "#333333") +
  geom_point(size = 4, color = "#2196F3") +
  scale_y_discrete(expand = expansion(mult = c(0.3, 0.3))) +
  theme_minimal() +
  labs(
    title = "Predictors of Remaining Undiagnosed with Hypertension",
    x = "Odds Ratio (OR > 1 means higher risk of being undiagnosed)",
    y = "Predictor Variable"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11, face = "bold"),
    axis.title.y = element_text(size = 11, face = "bold", margin = margin(r = 15))
  )

```

---

### 3. Model Interpretation

Based on the survey-weighted logistic regression, I observed the following within the hypertensive subpopulation:

* **Age:** There is a statistically significant negative association between age and the odds of remaining undiagnosed (**OR = 0.963, p < 0.001**). As age increases, patients are significantly less likely to remain undiagnosed, likely due to more frequent interactions with the healthcare system.
* **Income Ratio:** Higher income shows a slight, marginally significant positive association with remaining undiagnosed (**OR = 1.08, p = 0.0498**).
* **Gender:** There is no statistically significant difference in diagnosis rates between females and males in this cohort (**OR = 1.13, p = 0.428**), as the 95% confidence interval widely crosses 1.0.

---

## Conclusion & Health Policy Implications

* **Screening Focus:** Outreach campaigns should target young and middle-aged adults (ages 18–59), who exhibit high proportions of undiagnosed hypertension despite lower overall prevalence.
* **Methodological Rigor:** Standard non-survey analysis would misestimate standard errors; maintaining full PSU/Strata architecture via domain filtering (`srvyr::filter`) was essential to ensuring mathematically valid statistical inferences.

```